Extrinsic Type Inference (DRAFT 2025-06-09)

### 2025-07-14 extrinsic_type_inference_paper/paper.tex
update diff case and negatable and subtractable checks

\begin{definition}
  \label{def:type}
  \small
  \nopad
  \[\begin{array}[t]{rr@{}l}
    \com{type} &
    \nonterm{\tau} & \is 
    \begin{array}[t]{@{}l}
      \alpha \sep
      \J{@} \sep
      \kappa \sep 
      \phi \sep 
      \psi \sep 
      \J{LFP[}\alpha\J{]}\tau \sep
      \tau \typdiff \eta \sep
      \gamma
    \end{array}
    \\
    \com{implication} &
    \nonterm{\kappa} & \is 
    \begin{array}[t]{@{}l}
      \J{<} l \J{>} \tau \sep 
      \tau\J{->}\tau
    \end{array}
    \\
    \com{expansion} &
    \nonterm{\phi} & \is 
    \begin{array}[t]{@{}l}
      \tau\J{|}\tau \sep 
      \J{EXI[}\Theta\J{]} \Delta \J{:} \tau
    \end{array}
    \\
    \com{refinement} &
    \nonterm{\psi} & \is 
    \begin{array}[t]{@{}l}
      \tau\J{\&}\tau \sep 
      \J{ALL[}\Theta\J{]} \Delta \J{:} \tau
    \end{array}
    \\
    \com{negatable} &
    \nonterm{\eta} & \is 
    \begin{array}[t]{@{}l}
      \J{EXI[}\Theta \J{]} \Omega \J{:} \eta \sep 
      \rho
    \end{array}
    \\
    \com{pattern} &
    \nonterm{\rho} & \is 
    \begin{array}[t]{@{}l}
      \alpha \sep
      \J{@} \sep
      \J{<}l\J{>}\rho \sep 
      \rho\J{\&}\rho \sep
      \rho\J{*}\rho
    \end{array}
    \\
    \com{sugar} &
    \nonterm{\gamma} & \is 
    \begin{array}[t]{@{}l}
      \J{TOP} \sep \J{BOT} \sep 
      \tau \J{*} \tau 
    \end{array}
  \end{array}\]
\end{definition}

@@

A least fixed point \ms{\J{LFP[}\alpha \J{]}}
represents an infinite space of structures, 
such as natural numbers or lists. 
A difference \ms{\tau \typdiff \eta} is a space without inhabitation by certain structures.
The subtracted part is restricted to a negatable form,
which is a form that our system can decide the inhabitation
assuming additional semantics restrictions.

@@

\begin{definition}
  \label{def:sequence}
  \small
  \nopad
  \[\begin{array}[t]{rr@{}l}
    \com{type var seq} &
    \nonterm{\Theta} & \is \epsilon \sep \Theta\ \alpha 
    \\
    \com{subtyping seq} &
    \nonterm{\Delta} & \is \epsilon \sep \Delta \  \tau \J{<:} \tau
    \\
    \com{typing seq} &
    \nonterm{\Gamma} & \is \epsilon \sep \Gamma\ x \hastype \tau
    \\
    \com{schema seq} &
    \nonterm{\Pi} & \is \epsilon \sep \Pi\ (\Theta, \Delta, \tau) 
    \\
    \com{negatable type seq} &
    \nonterm{\Xi} & \is \epsilon \sep \Xi\ \eta 
    \\
    \com{negatable subtyping seq} &
    \nonterm{\Omega} & \is \epsilon \sep 
      \Omega\ \tau \J{<:} \eta \sep 
      \Omega\ \alpha \J{<:} \tau
    \\
    \com{composite seq} &
    \nonterm{C} & \is \epsilon \sep C\ \iota 
    \\
    \com{item} &
    \nonterm{\iota} & \is 
    \alpha
    \sep 
    \tau \J{<:} \tau
    \sep
    x \hastype \tau
    \sep
    (\Theta, \Delta, \tau)
    \sep
    \eta
    \sep
    \tau \J{<:} \eta

@@

For a function \ms{f},  
it delegates the work to two helpers. 
First, it leverages function typing consolidation 
\ms{\Theta, \Delta, \Gamma \entails f \liftfun \Pi \sim \Xi}
to infer as many types as possible
and consolidates the results into a sequence of schemas \ms{\Pi}.
It must consolidate all of the inferred types in order to
ensure that the inferred type of each 
path's pattern does not subtype the type of subsequent paths' patterns. 
The sequence of negatable types \ms{\Xi} can be ignored, as it is only
exposed for the sake of being used recursively.
Next, it leverage positive packing 
\ms{\textbf{pack}^+(... \entails \Pi) = \tau}
to rewrite each schema into a type,
such that the skolems and plastics are converted to
existential and universal bindings,
and then combine all the types by compounding intersection.
Packing the path types together ensures that certain
control flows of application can be verified even when  
the best approximation of the type of the argument
is less precise than the pattern of any one path.

@@

\begin{definition} 
  \label{def:proof_subtyping_difference_intro}
  \emph{proof subtyping (difference introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \typdiff \eta \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \Vvdash \eta
      \\
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \given \Theta', \Delta' 
      \\
      \nexists \Theta', \Delta' \qua  
      \Theta, \Delta \entails 
      \tau_l \subtypes \eta \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \typdiff \eta
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

\noindent
For difference introduction \ms{\tau \subtypes \tau_r \typdiff \eta}, 
it checks that subtyping holds for the positive part 
and fails for the subtracted part.
Additionally, it requires that the subtracted part is negatable
\ms{\Theta, \Delta \Vvdash \eta}, 
which ensures that its introduction is complete,
thus failure of its introduction is sound.

>>>>

\begin{definition}
  \label{def:type}
  \small
  \nopad
  \[\begin{array}[t]{rr@{}l}
    \com{type} &
    \nonterm{\tau} & \is 
    \begin{array}[t]{@{}l}
      \alpha \sep
      \J{@} \sep
      \kappa \sep 
      \phi \sep 
      \psi \sep 
      \J{LFP[}\alpha\J{]}\tau \sep
      \tau \typdiff \eta \sep
      \gamma
    \end{array}
    \\
    \com{implication} &
    \nonterm{\kappa} & \is 
    \begin{array}[t]{@{}l}
      \J{<} l \J{>} \tau \sep 
      \tau\J{->}\tau
    \end{array}
    \\
    \com{expansion} &
    \nonterm{\phi} & \is 
    \begin{array}[t]{@{}l}
      \tau\J{|}\tau \sep 
      \J{EXI[}\Theta\J{]} \Delta \J{:} \tau
    \end{array}
    \\
    \com{refinement} &
    \nonterm{\psi} & \is 
    \begin{array}[t]{@{}l}
      \tau\J{\&}\tau \sep 
      \J{ALL[}\Theta\J{]} \Delta \J{:} \tau
    \end{array}
    \\
    \com{subtractable} &
    \nonterm{\eta} & \is 
    \begin{array}[t]{@{}l}
      \J{EXI[}\Theta \J{]} \eta \sep 
      \rho
    \end{array}
    \\
    \com{pattern} &
    \nonterm{\rho} & \is 
    \begin{array}[t]{@{}l}
      \alpha \sep
      \J{@} \sep
      \J{<}l\J{>}\rho \sep 
      \rho\J{\&}\rho \sep
      \rho\J{*}\rho
    \end{array}
    \\
    \com{sugar} &
    \nonterm{\gamma} & \is 
    \begin{array}[t]{@{}l}
      \J{TOP} \sep \J{BOT} \sep 
      \tau \J{*} \tau 
    \end{array}
  \end{array}\]
\end{definition}

@@

A least fixed point \ms{\J{LFP[}\alpha \J{]}}
represents an infinite space of structures, 
such as natural numbers or lists. 
A difference \ms{\tau \typdiff \eta} is a space without inhabitation by certain structures.
The subtracted part is restricted to a subtractable form,
which is a form that our system can decide the inhabitation
assuming additional semantics restrictions.

@@

\begin{definition}
  \label{def:sequence}
  \small
  \nopad
  \[\begin{array}[t]{rr@{}l}
    \com{type var seq} &
    \nonterm{\Theta} & \is \epsilon \sep \Theta\ \alpha 
    \\
    \com{subtyping seq} &
    \nonterm{\Delta} & \is \epsilon \sep \Delta \  \tau \J{<:} \tau
    \\
    \com{typing seq} &
    \nonterm{\Gamma} & \is \epsilon \sep \Gamma\ x \hastype \tau
    \\
    \com{schema seq} &
    \nonterm{\Pi} & \is \epsilon \sep \Pi\ (\Theta, \Delta, \tau) 
    \\
    \com{subtractable type seq} &
    \nonterm{\Xi} & \is \epsilon \sep \Xi\ \eta 
    \\
    \com{negatable subtyping seq} &
    \nonterm{\Omega} & \is \epsilon \sep 
      \Omega\ \tau \J{<:} \eta \sep 
      \Omega\ \alpha \J{<:} \tau
    \\
    \com{composite seq} &
    \nonterm{C} & \is \epsilon \sep C\ \iota 
    \\
    \com{item} &
    \nonterm{\iota} & \is 
    \alpha
    \sep 
    \tau \J{<:} \tau
    \sep
    x \hastype \tau
    \sep
    (\Theta, \Delta, \tau)
    \sep
    \eta
    \sep
    \tau \J{<:} \eta

@@

For a function \ms{f},  
it delegates the work to two helpers. 
First, it leverages function typing consolidation 
\ms{\Theta, \Delta, \Gamma \entails f \liftfun \Pi \sim \Xi}
to infer as many types as possible
and consolidates the results into a sequence of schemas \ms{\Pi}.
It must consolidate all of the inferred types in order to
ensure that the inferred type of each 
path's pattern does not subtype the type of subsequent paths' patterns. 
The sequence of subtractable types \ms{\Xi} can be ignored, as it is only
exposed for the sake of being used recursively.
Next, it leverage positive packing 
\ms{\textbf{pack}^+(... \entails \Pi) = \tau}
to rewrite each schema into a type,
such that the skolems and plastics are converted to
existential and universal bindings,
and then combine all the types by compounding intersection.
Packing the path types together ensures that certain
control flows of application can be verified even when  
the best approximation of the type of the argument
is less precise than the pattern of any one path.

@@

\begin{definition} 
  \label{def:proof_subtyping_difference_intro}
  \emph{proof subtyping (difference introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \typdiff \eta \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \TODO{Update proof, deriving that lower does not intersect with subtraction}
      \\\\
      \Theta, \Delta \Vvdash \eta
      \\
      \nexists \Theta', \Delta' \qua  
      \Theta, \Delta \entails 
      \tau_l \subtypes \eta \given \Theta', \Delta'
      \\\\
      \Theta, \Delta \Vvdash \eta \subtypes  \tau_l
      \\
      \nexists \Theta', \Delta' \qua  
      \Theta, \Delta \entails 
      \eta \subtypes \tau_l \given \Theta', \Delta'
      \\\\
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \typdiff \eta
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

\noindent
For difference introduction \ms{\tau \subtypes \tau_r \typdiff \eta}, 
it checks that subtyping holds for the positive part 
and fails for the subtracted part.
Additionally, it requires that the subtracted part is subtractable 
\ms{\Theta, \Delta \Vvdash \eta}, 
which ensures that its introduction is complete,
thus failure of its introduction is sound.
